A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-05-18 inst/manuscript/manuscript.tex
minor formatting updates

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches in a finite population spatial context.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated and real
    data from the United States Environmental Protection Agency's 2012
    National Lakes Assessment. A variety of sample sizes, location
    layouts, dependence structures, and response types are considered. The
    population mean is the parameter of interest and performance is
    measured using statistics like bias, squared error, and interval
    coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the generalized random
    tessellation stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    simple random sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based inference tends to outperform design-based inference, even
    for skewed data where the model-based distributional assumptions are
    violated. The performance gap between design-based inference and
    model-based inference is small when GRTS samples are used but large
    when SRS samples are used, suggesting that the sampling choice
    (whether to use GRTS or SRS) is most important when performing
    design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial sampling and
    inference that practitioners must consider when choosing between them.
    We provide relevant background contextualizing each approach and study
    their properties in a variety of scenarios, making recommendations for
    use based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- see Ver Hoef (2008) for technical details and see Higham, J. Ver
Hoef, et al. (2021) for an extension to cases of imperfect detection
among population units. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes \(\bm{\delta}\) in Equation\(~\)\eqref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the separation vector (e.g., distance) between population units),
isotropic (independent of direction), and decays with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\). Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or not either. A variety of
flexible covariance functions can be used to model \(\bm{\delta}\)
(Cressie, 1993) -- one example is the exponential covariance function.
Cressie (1993) provides a thorough list of spatial covariance functions.
The \(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter that
quantifies the spatially dependent (correlated) variability,
\(\sigma^2_{2}\) is the variance parameter the quantifies that spatially
independent (not correlated) variability, \(\phi\) is the distance
parameter that measures the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). In geostatistical literature, \(\sigma^2_{1}\) is called the
partial sill, \(\sigma^2_{2}\) is called the nugget, and \(\phi\) is
called the range. We denote \(\bm{\theta}\) as the vector of covariance
parameters that composes \(\bm{\delta}\). In
Equation\(~\)\ref{equation:expcov},
\(\bm{\theta} = \{\sigma^2_1, \sigma^2_2, \phi \}\).

The parameters in Equation\(~\)\ref{equation:Zmarginal} can be estimated
using a variety of techniques, but we focus on using restricted maximum
likelihood (REML) (Harville, 1977; Patterson and Thompson, 1971;
Wolfinger et al., 1994). REML is preferred over maximum likelihood (ML)
because ML estimates can be badly biased for small sample sizes, due to
the fact that ML makes no adjustment for the simultaneous estimation of
\(\bm{\beta}\) and \(\bm{\theta}\) (Patterson and Thompson, 1971). Minus
twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation} where
\(\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s\)
and \(\bm{\Sigma}_{ss}\) is the covariance matrix of the sampled sites.
Minimizing Equation\(~\)\ref{equation:reml} yields
\(\bm{\hat{\theta}}_{reml}\), the REML estimates of \(\bm{\theta}\).
Then \(\bm{\hat{\beta}}_{reml}\), the REML estimate of \(\bm{\beta}\),
is given by
\((\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s\),
where \(\bm{\hat{\Sigma}}_{ss}\) is \(\bm{\Sigma}_{ss}\) evaluated at
\(\bm{\hat{\theta}}_{reml}\).

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate the variance. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the \(i\)th response value, \(\bar{z}\)
is the mean of all \(z_i\), \(n\) is the sample size, \(N\) is the
population size, and \(f = (1 - n / N)\) (\(f\) is often called the
finite population correction factor). For GRTS-DB, the Horvitz-Thompson
estimator was used to estimate means and the local neighborhood variance
was used to estimate variances. For SRS-MB and GRTS-MB, FPBK was used to
estimate means and variances using restricted maximum likelihood. SRS,
GRTS sampling, and design-based inference were implemented using the
\texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle et al., 2022).
FPBK was implemented using the \texttt{sptotal} \textbf{\textsf{R}}
package (Higham, Ver Hoef, et al., 2021).

We used simulated data to compare the sampling-inference combinations
across many realized populations from the same data-generating
stochastic process. We then used real data from the 2012 National Lakes
Assessment to compare the sampling-inference combinations within a
single realized population (which is typically the case in reality).
With the simulated data, we were in control of the data-generating
stochastic process and the random sampling process. With the real data,
we were in control of only the random sampling process.

@@

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/symm_pop_hist.jpeg}
  \caption{Histogram of a realized population for the normal response.}
  \label{fig:symm_pop_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/skew_pop_hist.jpeg}
  \caption{Histogram of a realized population for the skewed response.}
  \label{fig:skew_pop_hist}
\end{subfigure}
\caption{Histograms of realized populations simulated for the normal and skewed resposnes using the random layout and 50\% DRE.}
\label{fig:sim_pops}
\end{figure}

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. Within each simulation scenario and trial, SRS and
GRTS samples were selected and then design-based and model-based
inferences were used to estimate (design-based) or predict (model-based)
the realized mean and construct 95\% confidence (design-based) or 95\%
prediction (model-based) intervals. With model-based inference,
covariance parameters and the realized mean were estimated (using REML)
separately for each trial. After all 2000 trials, we summarized the
long-run performance of the sampling-inference combination in each
scenario by calculating mean bias, root-mean-squared error, and interval
coverage. Mean bias is taken as the average deviation between each
trial's estimated (or predicted) mean (\(\hat{\mu}_i\)) and its realized
mean (\(\mu_i\)):
\(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)\), where \(i\)
indexes simulation trials. Root-mean-squared error is taken as the
square root of the average squared deviation between each trial's
estimated (or predicted) mean and its realized mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}\).
Interval coverage is taken as the proportion of simulation trials where
the realized mean is contained in its 95\% confidence (or prediction)
interval. These intervals are constructed using the normal distribution
-- justification comes from the asymptotic normality of means via the
central limit theorem (under some assumptions). Quantifying these
metrics is important because together, they give us an idea of the
accuracy (mean bias), spread (RMSE), and validity (interval coverage) of
the sampling-inference combinations.

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the conterminous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within some distance class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance
increases steadily with distance (meaning that observations nearby one
another tend to be more similar than observations far apart from one
another).

@@

We studied performance of the four sampling-inference combinations by
selecting 2000 random SRS and GRTS samples of size \(n = 50\),
\(n = 100\), and \(n = 200\) from the realized ZMMI and Hg ppb
populations and then analyzing the samples using MB and DB inference. In
total, there were six separate scenarios (two responses crossed with
three sample sizes). Within each SRS and GRTS sample, design-based and
model-based inferences were used to estimate or predict the population
mean and construct 95\% coverage intervals. With model-based inference,
the exponential covariance was assumed, and covariance parameters and
the population mean were estimated using REML (separately for each SRS
and GRTS sample). We used the same evaluation metrics as for the
simulated data: mean bias, RMSE, and interval coverage. Mean bias is
taken as the average deviation between each sample's estimated (or
predicted) mean (\(\hat{\mu}_i\)) and the population mean (\(\mu\)) (of
ZMMI or Hg ppb): \(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)\),
where \(i\) indexes simulation trials. Root-mean-squared error is taken
as the square root of the average squared deviation between each
sample's estimated (or predicted) mean and its population mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}\).
Interval coverage is taken as the proportion of simulation trials where
the population mean is contained in its 95\% confidence (or prediction)
interval. These intervals are constructed using the normal distribution.

@@

95\% interval coverage of both ZMMI and Hg ppb for all four
sampling-inference combinations is shown in Fig. \ref{fig:data_figconf}.
When \(n = 50\), interval coverage for both responses is too low, though
interval coverage is higher for ZMMI (symmetric response) than for Hg
ppb (skewed response). When \(n = 100\), ZMMI interval coverage is
approximately 95\% except for GRTS-DB, which has coverage around 92\%,
while Hg ppb interval coverage ranges from approximately 90\% (GRTS-DB)
to 93\% (GRTS-MB). When \(n = 200\), ZMMI interval coverage is
approximately 95\% while Hg ppb interval coverage ranges from
approximately 93\% (GRTS-DB) to 95\% (GRTS-MB). As with the simulated
data, coverages for the NLA data tend to increase with the sample sizes,
coverages tend to be higher for symmetric responses than for skewed
responses, and the local neighborhood variance was slightly too small
for small \(n\), yielding slightly lower interval coverages than the
other sampling-inference combinations. Recall that model-based inference
defines interval coverage properties across realized populations. With
the simulated data, we evaluated interval coverage across realized
populations, but for the NLA data, we evaluated interval coverage within
a single realized population for different samples. We did find that
model-based coverages were similar to the design-based coverages,
however, suggesting that for some realized populations it is reasonable
to heuristically view data from separate random samples as being from
approximately separate realized populations. But generally, if
model-based intervals constructed from many random samples of a single
realized population show improper coverage, this does not necessarily
imply a deficiency in model-based inference. Tables for interval
coverage in all six simulation scenarios are provided in the supporting
information.

@@

The design-based and model-based approaches to frequentist statistical
inference rest on fundamentally different foundations. Design-based
approaches rely on random sampling to estimate population parameters.
Model-based approaches rely on distributional assumptions to predict
realized values of a data-generating stochastic process. Though
model-based approaches do not rely on random sampling, it can still be
beneficial as a way to guard against preferential sampling. While
design-based and model-based approaches have often been compared in the
literature from theoretical and analytical perspectives, our
contribution lies in studying them for finite population spatial data
while implementing GRTS sampling and the local neighborhood variance
estimator. Aside from the theoretical differences described throughout
the manuscript, a few analytical findings from the simulated and real
data studies were particularly notable. All sampling-inference
combinations had approximately zero mean bias. Independent of the
inference approach, GRTS-DB and GRTS-MB had lower RMSE than their SRS
counterparts. Though GRTS-DB and GRTS-MB generally had very similar
RMSE, SRS-MB tended to have much lower RMSE than SRS-DB, suggesting that
the model-based inference mitigated much of the inefficiency in RMSE
from SRS. As the proportion of dependent random error in the simulated
data increased, SRS-MB, GRTS-DB, and GRTS-MB become increasingly more
efficient (lower RMSE) than SRS-DB. Interval coverage tended to be
higher for the symmetric responses than skewed responses and tended to
increase with the sample size. At a sample size of \(n = 200\),
generally all interval coverages were near the desired value of 95\%.

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches in a finite population spatial context.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated data and
    real data. The real data is from the United States Environmental
    Protection Agency's 2012 National Lakes Assessment. A variety of
    sample sizes, location layouts, dependence structures, and response
    types are considered. The population mean is the parameter of
    interest, and performance is measured using statistics like bias,
    squared error, and interval coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the generalized random
    tessellation stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    simple random sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based inference tends to outperform design-based inference, even
    for skewed data where the model-based distributional assumptions are
    violated. The performance gap between design-based inference and
    model-based inference is small when GRTS samples are used but large
    when SRS samples are used, suggesting that the sampling choice
    (whether to use GRTS or SRS) is most important when performing
    design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial sampling and
    inference that practitioners must consider when choosing between them.
    We provide relevant background contextualizing each approach and study
    their properties in a variety of scenarios, making recommendations for
    use based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- see Ver Hoef (2008) for technical details and see Higham, Ver
Hoef, Madsen, et al. (2021) for an extension to cases of imperfect
detection among population units. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes \(\bm{\delta}\) in Equation\(~\)\eqref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the separation vector (e.g., distance) between population units), and
isotropic (independent of direction) (Cressie, 1993). Henceforth, it is
implied that we have made these same assumptions regarding
\(\bm{\delta}\). Chiles and Delfiner (1999), pp.~80-93 discuss
covariance functions that are not second-order stationary, not
isotropic, or not either. A variety of flexible covariance functions can
be used to model \(\bm{\delta}\) (Cressie, 1993) -- one example is the
exponential covariance function. Cressie (1993) provides a thorough list
of spatial covariance functions. The \(i,j\)th element of the
exponential covariance matrix, \(\mathop{\mathrm{{cov}}}(\bm{\delta})\),
is \mbox{} \begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter that
quantifies the spatially dependent (correlated) variability,
\(\sigma^2_{2}\) is the variance parameter the quantifies that spatially
independent (not correlated) variability, \(\phi\) is the distance
parameter that measures the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). In geostatistical literature, \(\sigma^2_{1}\) is called the
partial sill, \(\sigma^2_{2}\) is called the nugget, and \(\phi\) is
called the range. We denote \(\bm{\theta}\) as the vector of covariance
parameters that composes \(\bm{\delta}\). In
Equation\(~\)\ref{equation:expcov},
\(\bm{\theta} = \{\sigma^2_1, \sigma^2_2, \phi \}\).

The parameters in Equation\(~\)\ref{equation:Zmarginal} can be estimated
using a variety of techniques, but we focus on restricted maximum
likelihood (REML) (Harville, 1977; Patterson and Thompson, 1971;
Wolfinger et al., 1994). REML is preferred over maximum likelihood (ML)
because ML estimates can be badly biased for small sample sizes, due to
the fact that ML makes no adjustment for the simultaneous estimation of
\(\bm{\beta}\) and \(\bm{\theta}\) (Patterson and Thompson, 1971). Minus
twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation} where
\(\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s\)
and \(\bm{\Sigma}_{ss}\) is the covariance matrix of the sampled sites.
Minimizing Equation\(~\)\ref{equation:reml} yields
\(\bm{\hat{\theta}}_{reml}\), the REML estimates of \(\bm{\theta}\).
Then \(\bm{\hat{\beta}}_{reml}\), the REML estimate of \(\bm{\beta}\),
is given by
\((\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s\),
where \(\bm{\hat{\Sigma}}_{ss}\) is \(\bm{\Sigma}_{ss}\) evaluated at
\(\bm{\hat{\theta}}_{reml}\).

@@

In this section we describe how we used simulated and real data to
investigate performance between simple random sampling (SRS) and GRTS
sampling as well as performance between design-based (DB) and
model-based (MB) inference. In SRS and GRTS sampling, all population
units had equal inclusion probabilities and were selected without
replacement. The important distinction between SRS and GRTS is that SRS
ignores spatial locations while sampling but GRTS explicitly
incorporates them. Together, the two sampling plans (SRS and GRTS)
combined with the two inference approaches (DB and MB) yielded four
sampling-inference combinations: SRS-DB, SRS-MB, GRTS-DB, and GRTS-MB.
For SRS-DB, the Horvitz-Thompson estimator \eqref{eq:ht} was used to
estimate means and the commonly-used SRS variance formula (Lohr, 2009;
Särndal et al., 2003) was used to estimate variances. This variance
formula is given by \begin{equation}\label{equation:srs_var}
 \frac{f[\sum_{i = 1}^n (z_i - \bar{z})^2]}{n(n - 1)},
\end{equation} where \(z_i\) is the response value of the \(i\)th
sampled population unit, \(\bar{z}\) is the mean of all \(z_i\), \(n\)
is the sample size, \(N\) is the population size, and
\(f = (1 - n / N)\) (\(f\) is often called the finite population
correction factor). For GRTS-DB, the Horvitz-Thompson estimator was used
to estimate means and the local neighborhood variance was used to
estimate variances. For SRS-MB and GRTS-MB, FPBK was used to estimate
means and variances using restricted maximum likelihood. SRS, GRTS
sampling, and design-based inference were implemented using the
\texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle et al., 2022).
FPBK was implemented using the \texttt{sptotal} \textbf{\textsf{R}}
package (Higham, Ver Hoef, Frank, et al., 2021).

The simulated and real data were used in distinct ways to answer
distinct questions. We used the simulated data to compare the
sampling-inference combinations across many realized populations (from
the same data-generating stochastic process) and random samples. We used
the real data to compare the sampling-inference combinations within a
single realized population but across random samples. With the simulated
data, we were in control of the data-generating stochastic process and
the random sampling process. With the real data, we were only in control
of the random sampling process, which is typically the case in reality.

@@

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/symm_pop_hist.jpeg}
  \caption{Histogram of a single realized population for the normal response.}
  \label{fig:symm_pop_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/skew_pop_hist.jpeg}
  \caption{Histogram of a single realized population for the skewed response.}
  \label{fig:skew_pop_hist}
\end{subfigure}
\caption{Histograms of single realized populations simulated for the normal and skewed respones using the random layout and 50\% DRE.}
\label{fig:sim_pops}
\end{figure}

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. Within each trial, a population was simulated
according to the specifications of the particular simulation scenario
(for the random location layout, locations were simulated separately for
each trial). Next, SRS and GRTS samples were selected. Then,
design-based and model-based inferences were used to estimate
(design-based) or predict (model-based) the realized mean and construct
95\% confidence (design-based) or 95\% prediction (model-based)
intervals. With model-based inference, covariance parameters and the
realized mean were estimated (using REML) separately for each trial.
After all 2000 trials, we summarized the long-run performance of the
sampling-inference combination in each scenario by calculating mean
bias, root-mean-squared error, and interval coverage. Mean bias was
taken as the average deviation between each trial's estimated (or
predicted) mean (\(\hat{\mu}_i\)) and its realized mean (\(\mu_i\)):
\(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)\), where \(i\)
indexes the simulation trials. Because each trial had different realized
populations, \(\mu_i\) changed with \(i\). Root-mean-squared error was
taken as the square root of the average squared deviation between each
trial's estimated (or predicted) mean and its realized mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu_i)^2}\).
Interval coverage was taken as the proportion of simulation trials where
the realized mean was contained in its 95\% confidence (or prediction)
interval. These intervals were constructed using the normal distribution
-- justification comes from the asymptotic normality of means via the
central limit theorem (under some assumptions). Quantifying these
metrics is important because together, they give us an idea of the
accuracy (mean bias), spread (RMSE), and validity (interval coverage) of
the sampling-inference combinations.

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the conterminous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within a separation class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance
increases steadily with distance (meaning that observations nearby one
another tend to be more similar than observations far apart from one
another).

@@

We studied performance of the four sampling-inference combinations by
selecting 2000 SRS and GRTS samples of size \(n = 50\), \(n = 100\), and
\(n = 200\) from the realized ZMMI and Hg ppb populations and then
analyzing the samples using MB and DB inference. In total, there were
six separate scenarios (two responses crossed with three sample sizes).
Within each SRS and GRTS sample, design-based and model-based inferences
were used to estimate or predict the population mean and construct 95\%
coverage intervals. With model-based inference, the exponential
covariance was assumed, and covariance parameters and the population
mean were estimated using REML (separately for each SRS and GRTS
sample). We used the same evaluation metrics as for the simulated data:
mean bias, RMSE, and interval coverage. Mean bias was taken as the
average deviation between each sample's estimated (or predicted) mean
(\(\hat{\mu}_i\)) and the population mean (\(\mu\)) (of ZMMI or Hg ppb):
\(\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)\), where \(i\)
indexes the simulation trials. Because each trial had the same realized
population, \(\mu\) did not change with \(i\) (in contrast to the
simulated data, where the realized mean changed with \(i\)).
Root-mean-squared error was taken as the square root of the average
squared deviation between each sample's estimated (or predicted) mean
and its population mean:
\(\sqrt{\frac{1}{n}\sum_{i = 1}^{2000} (\hat{\mu}_i - \mu)^2}\).
Interval coverage was taken as the proportion of simulation trials where
the population mean was contained in its 95\% confidence (or prediction)
interval. These intervals were constructed using the normal
distribution.

@@

95\% interval coverage of both ZMMI and Hg ppb for all four
sampling-inference combinations is shown in Fig. \ref{fig:data_figconf}.
When \(n = 50\), interval coverage for both responses is too low, though
interval coverage is higher for ZMMI (symmetric response) than for Hg
ppb (skewed response). When \(n = 100\), ZMMI interval coverage is
approximately 95\% except for GRTS-DB, which has coverage around 92\%,
while Hg ppb interval coverage ranges from approximately 90\% (GRTS-DB)
to 93\% (GRTS-MB). When \(n = 200\), ZMMI interval coverage is
approximately 95\% while Hg ppb interval coverage ranges from
approximately 93\% (GRTS-DB) to 95\% (GRTS-MB). As with the simulated
data, coverages for the NLA data tend to increase with the sample sizes,
coverages tend to be higher for symmetric responses than for skewed
responses, and the local neighborhood variance was slightly too small
for small \(n\), yielding slightly lower interval coverages than the
other sampling-inference combinations. Recall that model-based inference
defines interval coverage properties across realized populations. With
the simulated data, we evaluated interval coverage across realized
populations, but for the NLA data, we evaluated interval coverage within
a single realized population. We did find that model-based coverages
were similar to the design-based coverages, however, suggesting that for
some realized populations it is reasonable to heuristically view data
from separate random samples as being from approximately separate
realized populations. But generally, if model-based intervals
constructed from many random samples of a single realized population
show improper coverage, this does not necessarily imply a deficiency in
model-based inference. Tables for interval coverage in all six
simulation scenarios are provided in the supporting information.

@@

The design-based and model-based approaches to frequentist statistical
inference rest on fundamentally different foundations. Design-based
approaches rely on random sampling to estimate population parameters.
Model-based approaches rely on distributional assumptions to predict
realized values of a data-generating stochastic process. Though
model-based approaches do not rely on random sampling, random sampling
can still be beneficial as a way to guard against preferential sampling.
While design-based and model-based approaches have often been compared
in the literature from theoretical and analytical perspectives, our
contribution lies in studying them for finite population spatial data
while implementing GRTS sampling and the local neighborhood variance
estimator. Aside from the theoretical differences described throughout
the manuscript, a few analytical findings from the simulated and real
data studies were particularly notable. All sampling-inference
combinations had approximately zero mean bias. Independent of the
inference approach, the GRTS samples yielded lower RMSE than their SRS
counterparts. Though GRTS-DB and GRTS-MB generally had very similar
RMSE, SRS-MB tended to have much lower RMSE than SRS-DB, suggesting that
the model-based inference mitigated much of the inefficiency in RMSE
from SRS. As the proportion of dependent random error in the simulated
data increased, SRS-MB, GRTS-DB, and GRTS-MB become increasingly more
efficient (lower RMSE) than SRS-DB. Interval coverage tended to be
higher for the symmetric responses than skewed responses and tended to
increase with the sample size. At a sample size of \(n = 200\),
generally all interval coverages were near the desired value of 95\%.
